Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-29 paper/paper.Rmd
updating

---
title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
shorttitle        : "Ordinal regression models made easy"
author: 
  - name          : "Filippo Gambarota"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Via Venezia 8, 35131, Padova (Italy)"
    email         : "filippo.gambarota@unipd.it"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Methodology"
      - "Software"
  - name          : "Gianmarco Altoè"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
abstract: |
  Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables. Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression). This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction. One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis. The aim of the tutorial is presenting ordinal regression models using a simulation-based approach. In the first section we introduced the general model highlighting the crucial components and assumptions. Then we explained how to intepret the parameters for a logit and probit model. Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.

@@

In the first part we introduce the ordinal models explaining the general structure and assumptions. Then we move with model fitting and parameters interpretation. Finally we introduce the simulation approach for common research scenarios and an application to power analysis. We used `r cite_r()` for the code, figures and tables. Details about R packages an The tutorial is supported by a set of custom R functions available on Github. Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials). For the tutorial we used the following packages.

@@

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials contains some examples of these complex models. Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

>>>>

---
title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
shorttitle        : "Ordinal regression models made easy"
author: 
  - name          : "Filippo Gambarota"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Via Venezia 8, 35131, Padova (Italy)"
    email         : "filippo.gambarota@unipd.it"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Methodology"
      - "Software"
  - name          : "Gianmarco Altoè"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
abstract: |
  Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables. Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression). This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction. One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis. The aim of the tutorial is presenting ordinal regression models using a simulation-based approach. In the first section we introduced the general model highlighting the crucial components and assumptions. Then we explained how to intepret the parameters for a logit and probit model. Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).

@@

In the first part we introduce the ordinal models explaining the general structure and assumptions. Then we move with model fitting and parameters interpretation. Finally we introduce the simulation approach for common research scenarios and an application to power analysis. We used `r cite_r()` for the code, figures and tables. Details about R packages an The tutorial is supported by a set of custom R functions that can be found on the online OSF repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)). For the tutorial we used the following packages.

@@

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)) contains some examples of these complex models. Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
